Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2025-01-28 paper/paper.tex
paper

\section{Introduction}
\label{sec:introduction}
Modern CPUs are often equipped with multiple different vector registers. These registers are nowadays 
as wide as 512 bits, allowing for the processing of multiple data streams at once (SIMD). By 
batching multiple values together in one register, different ISAs like Intel AVX or ARM SVE allow 
the execution of the same instruction for all values in a vector register simultaneously. 
Utilizing this leads to significant performance improvements over the scalar equivalent most of 
the time.

@@

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, inner, and nested loops with complex 
control-flow.

The auto-vectorization pass would create multiple different VPlans based on initial legality 
checks. Each VPlan represents a different vectorization approach, by modeling the
underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
transformations can then optimize these abstracted control-flows based on a cost model.
After that, the best VPlan, and with it, the best vectorization approach is chosen by the cost model.
This VPlan is then executed by going over its own vectorized control flow representation and
materializing it back into the underlying LLVM IR.

LLVM's VPlan is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} feature 
flag~\cite{llvmouterloop,llvmouterloopstatus}. 
In the future, the plan will be to merge both of these loop vectorization paths into one. 
Merging the existing SLP Vectorizer with VPlan is also planned.

The rest of this paper is organized as follows. First, we explain different vectorization
strategies and constraints in general and describe the original auto-vectorization
approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan approach, focusing on
its most important individual components, and gives an outlook on future developments. In 
Section~\ref{sec:relatedwork}, we highlight work related to LLVM's auto-vectorization. Concluding,
we summarize VPlan, its current state, and its potential future.

@@

\paragraph{Inner Loop Vectorization}
Innermost loops are loops whose body's control flow does not contain any more loops.
Vectorization of these loops is more straightforward as the only option to vectorize is over 
the loop's induction variable.

@@

\paragraph{Outer Loop Vectorizaton}
Outer loop vectorization focuses on control flow with nested loops. Such control flow often presents
more challenges for auto-vectorization, as one could vectorize over the outer loop,
inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach.
Finding the best legal vectorization can therefor be more difficult.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.39\textwidth]{images/outer-loop-vec.png}
  \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
  outer loop. The inner loop has a trip count (6) that is too low to vectorize. Additionally,
  the memory access pattern of \texttt{y} would be consecutive compared to scattered when
  vectorizing over \texttt{i} instead of \texttt{j}.}
  \label{fig:outer-loop-vec}
\end{figure}

\paragraph{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control-flow but operating on multiple values at once through vector registers.

@@

Some of the most common vectorization barriers are data dependencies. As the vectorized version
of a code executes for Vectorization Factor (VF) $N$ data streams simultaneously, it inherently changes
the order of operations. To guarantee that the semantics of the code do not change, all
operations on these $N$ data streams must, generally, be independent. The simplest case
where this is not provided is when the $i$-th loop iteration depends on any of the $(i-N)$-th 
previous iterations.

@@

Uncountable loops, meaning loops where the total iteration count cannot be determined before the 
loop is executed, can often not be vectorized. This is because their exit condition usually depends on 
the loop's body itself, making it impossible to predetermine how many vector lanes are needed for the next 
loop iteration when executing.

Finally, some instructions do generally not have a direct vectorized 
instruction equivalent. For example Integer division, commonly implemented using the 
\textit{Newton-Raphson method}, is already a complex and iterative operation 
on modern CPUs. Applying this algorithm 
to multiple data elements in a single vector register would result in even more implementation complexity 
and is thus considered not worthwhile in most cases.
Workarounds for some specialized cases do exist~\cite{vecintdiv}.

\subsection{Vectorization Costs}
While vectorizing a given control flow may not be legal, the vectorized code sometimes performs 
worse than its scalar equivalent.

Some control flow transformations that seem beneficial at compile time might reveal a performance
bottleneck at runtime. Considering the code snippet from Figure~\ref{fig:inner-loop-vec}, it might
be that almost all \texttt{z[i]} evaluate to false at runtime. The branch predictor of a CPU
would then be able to guess the correct branch in the scalar code. The vectorized version would
always load the \texttt{y[i]} into memory only for the addition to be masked away. Different 
approaches to avoid such runtime penalties are branch-on-superword-condition code 
(BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
consolidation (ALC)~\cite{10.1007/s11227-022-04359-w,10.5555/3615924.3615932}.

The actual performance of a vector instruction also heavily depends on the ISA and its 
hardware implementation. 
For example, gather, scatter, 
shuffle, and permutation instructions are typically slow. Control flow with many 
non-adjacent memory accesses or 
operations that combine different vector elements from the same register (horizontal reductions) can thus 
decrease vectorization performance.

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the Vectorization Factor, a scalar
epilogue/tail of the loop might be used to handle the remaining loop iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop must be executed due to 
unclear data dependencies.

@@

However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
and nested loop vectorizations would have been difficult, as LV was designed to flatten a
loop's control flow into a single basic block. Producing and considering more and 
more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
not scale well.

As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
using a more modular and flexible system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

@@

Intel's initial VPlan proposal aimed to unite any 
auto-vectorization systems (SLP, Inner/Outer-Loop, Whole-Function) into a 
single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlans to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
optimization decisions by performing transformations on this CFG without modifying any LLVM IR.
Each VPlan could to calculate its cost and execute itself, materializing the vectorization 
approach back into LLVM IR.

@@

\subsection{VPlan Structure}
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

@@

\paragraph{H-CFG}
\emergencystretch 1em

This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling the vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
in that a single node of this graph can be a complete subgraph, representing more complex nested
behavior.
Specifically, any node of this H-CFG is one of three types:

\begin{itemize}
  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes} similar to 
  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
  \item \texttt{VPRegionBlock} represents a subgraph with a single VP entry node and a 
  single VP exit node. A \texttt{VPRegionBlock} can be marked to replicate its subgraph
  \texttt{VF * UF} times when executing. This is useful for modeling scalarized or predicated
  instructions within the final vectorized loop body.
  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
  IR \texttt{BasicBlock}. These nodes help model the scalar parts around the 
  final vectorized loops, such as the loop preheader and the epilogue loop.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop part of the hierarchical CFG of a VPlan modeling the 
  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
  Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the vectorized loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}. Note that the full H-CFG of this VPlan consists of many more nodes 
to handle legality checks (coming before the vectorized loop) or the entire scalar epilogue loop 
(coming after the vectorized loop).

The H-CFG in Figure~\ref{fig:inner-loop-vplan-hcfg-body} consists of 3 nodes on the highest layer. 
\texttt{vector.preheader} and \texttt{middle.block} 
are simple \texttt{VPBasicBlocks} surrounding the main vector loop node. This \texttt{vector loop} node 
is a \texttt{VPRegionBlock} that does 
not replicate its contents (indicated by \texttt{<x1>}; otherwise, it would be \texttt{<xVFxUF>}).

@@

\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions 
of the underlying scalar IR, called ingredients, and will materialize into one or more LLVM IR 
instructions when executed.

There are currently 40 different final types of \texttt{VPRecipes}. The types used 
within Figure~\ref{fig:inner-loop-vplan-hcfg-body} and the most important ones are:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
  \item \texttt{VPHeaderPHIRecipe} is a sub-type that models any induction, such as first-order 
  recurrence, pointer induction, or reduction as a phi node.
  \item \texttt{VPCanonicalIVPHIRecipe} derives from \texttt{VPHeaderPHIRecipe} and represents a 
  \texttt{CANONICAL-INDUCTION} variable via a \textit{scalar} value.
  \item \texttt{VPScalarIVStepsRecipe} models the individual offsets (\texttt{SCALAR-STEPS}) per 
  vector-lane from a common Integer or Floating-Point induction value.
\end{itemize}

Most of the \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have capitalized 
\texttt{HINTS} indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related in their semantical effect.

The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
This block decides if the scalar epilogue loop is executed after the vectorized loop by 
comparing the vector trip count to the original loop trip count. This additional vector trip count is 
needed as the original loop trip count might not be a perfect multiple of VF * UF, the number of elements 
processed in a single iteration.

The red recipes generally handle the induction variable \texttt{i} and the termination of the 
vectorized loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
\texttt{vp<\%1>}, the vector trip count. The per vector-lane offset's step size is 1 (\texttt{ir<1>}), 
meaning that each 
vector-lane offset from the base induction variable is one greater than the previous lane (we access the 
memory consecutively).

The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the mask, which 
results from this condition and is used for the guarded memory operations.

Finally, the green recipes represent the addition. First, the data from both \texttt{y} and \texttt{x} 
gets loaded (guarded by the compare mask), then added together, and then stored back into 
\texttt{x} (again, guarded by the compare mask).

@@

Such transformations include for example:
\begin{itemize}
  \item Simplifying \texttt{VPRecipes} by removing blend operations with only one unique input, removing 
  redundant type casts, and simplifying logical patterns.
  \item Removing dead \texttt{VPRecipes} whose values have no users.
  \item Moving loop-invariant \texttt{VPRecipes} out of the vector loop region. A \texttt{VPRecipe} is 
  considered loop-invariant if it does not have any side effects, does not read from memory, is not a 
  phi node, and all its operands are defined outside of the vector loop region.
\end{itemize}

@@

\subsection{Phase 1: Legality}
The main part of this first phase is checking whether it is even legal to vectorize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

Some of the most important legality checks done are:
\begin{itemize}
  \item The scalar loop's control flow must not contain indirect branches or multiple back-edges.
  An indirect branch is a branch that jumps to a value evaluated at runtime. A back-edge in a loop 
  is an edge from the loop back to it. 
  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
  statement must not exit the loop. All instructions of such a block can be executed speculatively 
  (e.g. via masking)
  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
  variable, or a fixed-order recurrence.
  \item The loop must not contain function calls outside of standard math library calls.
  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
\end{itemize}

Additionally, during this phase, all recurrences of the loop, such as reductions or inductions, are analyzed,
and their phi nodes are stored in a set.

\subsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop and it is deemed legal to vectorize the 
\texttt{LoopVectorizationCostModel} and the \texttt{LoopVectorizationPlanner} are initialized.
The latter one then constructs the initial VPlans based on the properties gathered during the legal phase
and the cost model initialization, such as the recurrence variables or the maximum viable vectorization 
factor.

A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

After the initial VPlans have been optimized, the best Vectorization Factor is computed. This is done 
by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
this step computes an optimal VF of $1$, any vectorization is not profitable compared to the 
scalar version.

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
architecture.

A remainder of the old \texttt{LoopVectorize} implementation before VPlan is the 
before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes different costs based on the scalar loop 
IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
after vectorization or the LLVM values for which the cost can 
be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
or trivially dead instructions.

Additionally, each \texttt{VPRecipe} has a function to calculate its own cost. For most \texttt{VPRecipes} 
that reference one or more LLVM IR ingredients, these are then used to calculate the cost of the recipe.
On the other hand, \texttt{VPRecipes} that do not directly model an underlying LLVM scalar IR instruction, 
such as \texttt{VPInstruction}, can generally not compute their own cost and thus return a cost of zero 
for now.

For simple innermost loops with few vectorization approaches,
relying on the old cost model might not have too much of a performance impact. 
However, once more VPlan transformations and complex
outer/hybrid loop vectorization are implemented, this new cost model would be an important
detail of the VPlan system to reliably choose the most optimal vectorization approach.

>>>>

\section{Introduction}
\label{sec:introduction}
Modern CPUs are often equipped with multiple different vector registers. These registers are nowadays 
as wide as 512 bits or more, allowing for the processing of multiple data streams at once (SIMD). By 
batching multiple values together in one register, different ISAs like Intel AVX or ARM SVE allow 
the execution of the same instruction for all values in a vector register simultaneously. 
Utilizing this leads to significant performance improvements over the scalar equivalent most of 
the time.

@@

With these limitations in mind, Intel started an ongoing refactorization effort to migrate LLVM's
auto-vectorization pipeline to utilize a more abstract Vectorization Plan 
(VPlan)~\cite{llvmextloopvec,llvmvplan}. The final goal would be to unite LLVM's auto-vectorization
in a single flexible system capable of optimizing SLPs, and inner and nested loops with complex 
control-flow.

The auto-vectorization pass would create multiple different VPlans based on initial legality 
checks. Each VPlan represents a different vectorization approach, by modeling the
underlying scalar control flow with its own abstracted vectorized version. Different VPlan-to-VPlan
transformations can then optimize these abstracted control-flows based on a cost model.
After that, the best VPlan, and with it, the best vectorization approach is chosen by the cost model.
This VPlan is then executed by going over its own vectorized intermediate representation and
materializing it back into the underlying LLVM IR.

LLVM's VPlan system is currently being integrated into the existing Loop Vectorizer and is
already modeling most inner loop vectorizations and transformations~\cite{llvmvplanupdate}. 
Vectorization for outer loops is also in development and can be enabled by setting 
the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} feature 
flag~\cite{llvmouterloop,llvmouterloopstatus}. 
In the future, the plan will be to merge both of these loop vectorization paths into one. 
Merging the existing SLP Vectorizer with VPlan is also planned.

The rest of this paper is organized as follows. First, we explain different vectorization
strategies and constraints in general and describe the original auto-vectorization
approach in LLVM. Section~\ref{sec:vplan} then introduces the new VPlan system, focusing on
its core VPlan component. Next, we summarize the current state of VPlan as of LLVM version 19.1.4.
Section~\ref{sec:vplanfuture} gives an outlook on future developments regarding VPlan. In 
Section~\ref{sec:relatedwork}, we highlight other approaches to auto-vectorization and 
comparisons of LLVM's auto-vectorization to other compilers such as GCC. 
Concluding, we summarize the intentions of VPlan, its current state, and its potential future.

@@

\paragraph{Inner Loop Vectorization}
Innermost loops are loops whose body's control flow does not contain any more loops.
Vectorization of these loops is more straightforward as the only option to vectorize is over 
the one loop's induction variables.

@@

\paragraph{Outer Loop Vectorizaton}
Outer loop vectorization focuses on control flow with nested loops. Such control flow often presents
more challenges for auto-vectorization, as one could vectorize over the outer loop,
inner loop, or a mix of both, representing a hybrid/multi-dimensional vectorization approach.
Finding the best legal vectorization can, therefore, be more difficult.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.39\textwidth]{images/outer-loop-vec.png}
  \caption{A code snippet of a nested loop where it would be beneficial to vectorize over the 
  outer loop. The inner loop has a trip count that is too low to vectorize (6). Additionally,
  the memory access pattern of \texttt{y} would be consecutive compared to scattered when
  vectorizing over \texttt{i} instead of \texttt{j}.}
  \label{fig:outer-loop-vec}
\end{figure}

\paragraph{Function Vectorization}
Function vectorization refers to the vectorization of entire functions. A loop might contain a
non-inlined function call, passing data related to the iteration variable. During auto-vectorization, 
this call could be replaced with a call to a newly constructed function with the same semantic 
control-flow but operating on multiple values at once by utilizing vector registers.

@@

Some of the most common vectorization barriers are data dependencies. As the vectorized version
of a code executes for Vectorization Factor (VF) data streams simultaneously, it inherently changes
the order of operations. To guarantee that the semantics of the code do not change, all
operations on these VF data streams must, generally, be independent. The simplest case
where this is not provided is when the $i$-th loop iteration depends on any of the 
previous iterations up to the $(i-\text{VF})$-th.

@@

Uncountable loops, meaning loops where the total iteration count cannot be determined before the 
loop is executed, can often not be vectorized. This is because their exit condition usually depends on 
the loop's body itself, making it impossible to predetermine how many iterations could be processed together 
before having to check the exit condition again.

Finally, some instructions do generally not have a direct vectorized 
instruction equivalent. For example Integer division, commonly implemented using the 
Newton-Raphson method, is already a complex and iterative operation 
on modern CPUs. Applying this algorithm 
to multiple data elements in a single vector register would result in even more implementation complexity 
and is thus considered not worthwhile in most cases.
Workarounds for some specialized cases do exist~\cite{vecintdiv}.

\subsection{Vectorization Costs}
While vectorizing a given control flow may not be legal, the vectorized code can sometimes also perform 
worse than its scalar equivalent.

Some control flow transformations that seem beneficial at compile time might reveal a performance
bottleneck at runtime. Considering the code snippet from Figure~\ref{fig:inner-loop-vec}, it might
be that almost all \texttt{z[i]} evaluate to false at runtime. The branch predictor of a CPU
would then be able to guess the correct branch in the scalar code. The vectorized version would
always access the memory at \texttt{y[i]} only for the addition to be masked away. Different 
approaches to avoid such runtime penalties are branch-on-superword-condition code 
(BOSCC)~\cite{10.5555/1299042.1299055,llvmboscc} or active-lane 
consolidation (ALC)~\cite{10.1007/s11227-022-04359-w,10.5555/3615924.3615932}.

The actual performance of a vector instruction also heavily depends on the ISA and its 
hardware implementation. 
For example, gather, scatter, 
shuffle, and permutation instructions are typically considered slow. Control flow with many 
non-adjacent memory accesses or 
operations that combine different vector elements from the same register (horizontal reductions) can thus 
decrease overall performance.

Additionally, the vectorized code size will always be larger than the scalar equivalent. This is
due to many factors: (1) The transformed control flow often includes more instructions.
(2) As the loop trip count might not be a perfect multiple of the vectorization factor, a scalar
epilogue/tail of the loop might be used to handle the remaining loop iterations. (3) Various runtime
checks might be inserted throughout the vectorized control flow to handle, for example, pointers 
to overlapping memory regions, in which case the scalar version of the loop must be executed.

@@

However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
and nested loop vectorizations would have been difficult, as LV was designed to flatten a
loop's control flow into a single basic block. Producing and considering more artifacts when trying to 
transform the scalar loop into a vectorized all at once also did not scale well.

As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
pipeline to use a more modular and scalable system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

@@

Intel's initial VPlan proposal aimed to unite any 
auto-vectorization strategy (SLP, Inner/Outer-Loop, Whole-Function) into a 
single flexible system~\cite{llvmextloopvec,llvmintrvplan,llvmvplanstate}. 
This system would use VPlan objects to represent different 
vectorization approaches, abstracting away from the underlying scalar loop IR using its own 
hierarchical control flow graph (H-CFG). This allows iteratively exploring different 
optimization decisions by performing transformations on this graph without modifying the 
underlying LLVM IR.
Each VPlan could calculate its cost and execute itself, materializing its vectorization 
approach back into the LLVM IR.

@@

\subsection{VPlan Structure}
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \textit{VPlan} in more detail.

@@

\paragraph{H-CFG}
\emergencystretch 2em

This is achieved by introducing a new layer of abstraction from the underlying loop's scalar IR.
Each VPlan stores its own Hierarchical-CFG (H-CFG) modeling its vectorization candidate. This 
control-flow graph and its components are the main elements of a single VPlan. It is hierarchical
in that a single node of this graph can be a complete subgraph, representing more complex nested
behavior.
Specifically, any node of this H-CFG is one of three types:

\begin{itemize}
  \item \texttt{VPBasicBlock} stores a list of \texttt{VPRecipes} similar to 
  the LLVM IR's \texttt{BasicBlock} holding a list of IR Instructions.
  \item \texttt{VPRegionBlock} represents a subgraph with a single entry and exit node. 
  A \texttt{VPRegionBlock} can be marked to replicate its subgraph
  \texttt{VF~*~UF} times when executing. This is useful for modeling scalarized or predicated
  instructions within the final vectorized loop body.
  \item \texttt{VPIRBasicBlock} is a special \texttt{VPBasicBlock} that wraps an underlying
  IR \texttt{BasicBlock}. These nodes help model the scalar parts around the 
  final vectorized loops, such as the loop preheader and the epilogue loop.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop part of a VPlan's hierarchical CFG modeling the 
  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
  Surrounding nodes were omitted for clarity.
  Some predefined values are
  \texttt{vp<\%0>} = \texttt{VF * UF}; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the equivalent vector loop part of a VPlan's H-CFG that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}. Note that the complete H-CFG of this VPlan consists of many more nodes 
to handle legality checks (coming before the vectorized loop) or the entire scalar epilogue loop 
(coming after the vectorized loop).

The H-CFG in Figure~\ref{fig:inner-loop-vplan-hcfg-body} consists of 3 nodes on the highest layer. 
\texttt{vector.preheader} and \texttt{middle.block} 
are simple \texttt{VPBasicBlocks} surrounding the main \texttt{vector loop} node. The \texttt{vector loop} node 
is a \texttt{VPRegionBlock} that does 
not replicate its contents (indicated by \texttt{<x1>}; otherwise, it would be \texttt{<xVFxUF>}).

@@

\texttt{VPRecipes} can be seen as instructions of an intermediate representation specialized for auto-vectorization. 
A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions 
of the underlying scalar IR. These underlying IR instructions are called \textit{ingredients}. 
Each \texttt{VPRecipe} will materialize back into one or more LLVM IR instructions when it is executed.

There are currently 40 different final types of \texttt{VPRecipes}. The types used 
within Figure~\ref{fig:inner-loop-vplan-hcfg-body} and the most important ones are:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or \texttt{VF * UF} times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
  \item \texttt{VPHeaderPHIRecipe} is a sub-type that models any induction, such as first-order 
  recurrence, pointer induction, or reduction as a phi node.
  \item \texttt{VPCanonicalIVPHIRecipe} derives from \texttt{VPHeaderPHIRecipe} and represents a 
  \texttt{CANONICAL-INDUCTION} variable via a \textit{scalar} value.
  \item \texttt{VPScalarIVStepsRecipe} models the individual offsets (\texttt{SCALAR-STEPS}) per 
  vector-lane from a common Integer or Floating-Point induction value.
\end{itemize}

Most of the \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have capitalized 
\texttt{HINTS} indicating their type. Additionally IR instructions/\texttt{VPRecipes} of the same color 
in Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related in their semantical effect.

The only \texttt{VPRecipes} in our example VPlan H-CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
This block decides if the scalar epilogue loop is still executed after the vectorized loop by 
comparing the vector trip count to the original loop trip count. This additional vector trip count is 
needed as the original loop trip count might not be a perfect multiple of \texttt{VF * UF}, the number of elements 
processed in a single vector loop iteration.

The red recipes generally handle the induction variable \texttt{i} and the termination of the 
vector loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
\texttt{vp<\%1>}, the vector trip count. The per vector-lane offset's step size is 1 (\texttt{ir<1>}), 
meaning that each 
vector-lane offset from the base induction variable (\texttt{vp<\%3>}) is one greater than the previous lane. 
Thus the memory is accessed consecutively.

The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the conditional mask 
and is used for the guarded memory operations.

Finally, the green recipes represent the addition. First, the data from both \texttt{y} and \texttt{x} 
gets loaded, then added together, and then stored back into \texttt{x}.

@@

Such transformations include for example:
\begin{itemize}
  \item Simplifying \texttt{VPRecipes} by removing blend operations with only one unique input, removing 
  redundant type casts, and simplifying logical patterns.
  \item Removing dead \texttt{VPRecipes} whose produced values have no users consuming them.
  \item Moving loop-invariant \texttt{VPRecipes} out of the vector loop region. A \texttt{VPRecipe} is 
  considered loop-invariant if it does not have any side effects, does not read from memory, is not a 
  phi node, and all its operands are defined outside of the vector loop region.
\end{itemize}

@@

\subsection{Phase 1: Legality}
The main part of this first phase is checking whether it is even legal to vectorize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LoopVectorize has a several extra conditions to check.

Some of the most important legality checks done are:
\begin{itemize}
  \item The scalar loop's control flow must not contain indirect branches or multiple back-edges.
  An indirect branch is a branch that jumps to a value evaluated at runtime. A back-edge in a loop 
  is an edge from the loop's body back to its start. 
  \item Loops with non-linear control flow must be if-convertible. This means any branch or switch 
  statement must not exit the loop. All instructions of such a predicated block can be executed speculatively 
  (e.g. via masking)
  \item Loop-carried dependencies must be vectorizable (see Section~\ref{sec:vec-legal}). In LV 
  this means explicitly that such a value must \textit{at least} either be an induction variable, a reduction 
  variable, or a fixed-order recurrence.
  \item The loop must not contain function calls outside of standard math library calls.
  \item The number of runtime checks to guarantee no pointers used by recurrences overlap must not be too high.
\end{itemize}

Additionally, during this phase, all recurrences of the loop, such as reductions or inductions, are analyzed,
and their phi nodes are stored for the later initialization of VPlans.

\subsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop and it was deemed legal to vectorize, the 
\texttt{LoopVectorizationCostModel} and \texttt{LoopVectorizationPlanner} are initialized.
The latter one then constructs the initial VPlans for different vectorization factors based on the properties 
gathered during the legal phase
and the cost model initialization. These properties are, for example, recurrence variables or the maximum viable 
vectorization factor.

A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar epilogue loop itself. Then the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After that the original scalar loop is represented by a general vectorization approach. Lastly, several VPlan 
transformations are executed, transforming and optimizing the VPlan's H-CFG and its \texttt{VPRecipes}.

After all the initial VPlans have been optimized, the best vectorization factor is computed. This is done 
by comparing all generated VPlans for all their possible VFs and choosing the most profitable one. If 
this step computes an optimal VF of $1$, any vectorization is not profitable compared to the 
scalar version.

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their direct vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
architecture.

A remainder of the old \texttt{LoopVectorize} implementation before VPlan is the 
before-mentioned \texttt{LoopVectorizationCostModel}. It is a single object that computes costs based on the scalar loop 
IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
after vectorization or the LLVM values for which the cost can 
be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
or trivially dead instructions.

Additionally, each \texttt{VPRecipe} has a function to calculate its own cost. For most \texttt{VPRecipes} 
that reference one or more LLVM IR ingredients, these are then used to calculate the cost of the recipe.
On the other hand, \texttt{VPRecipes} that do not directly model an underlying LLVM scalar IR instruction, 
such as \texttt{VPInstruction}, can generally not compute their own cost yet, thus returning a cost of zero 
for now.

For simple innermost loops with few vectorization approaches,
relying on the old cost model might not have too much of a performance impact. 
However, once more VPlan transformations and complex
outer/hybrid loop vectorization are implemented, this new cost model would be an important
detail of the VPlan system to reliably choose the most optimal vectorization candidate.

\subsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed.
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
Then a vectorized code skeleton is created in the IR, and the VPlan executed. 
The H-CFG of the VPlan is walked, and each node executed, which in turn executes all 
of its \texttt{VPRecipes} in order. Each \texttt{VPRecipe} produces its respective output IR.

A \texttt{VPTransformState} object is passed along to store information 
needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
used by later \texttt{VPRecipes}.
